### 2017-04-29 paper/paper.tex
add paper

\documentclass{article}

\usepackage[left=3cm,right=3cm,top=2cm,bottom=2cm]{geometry}
\usepackage[fleqn]{amsmath}
\usepackage{amssymb}
\usepackage{mathtools}
\usepackage{enumerate}
\usepackage{clrscode3e}
\usepackage{parskip}
\usepackage{mathptmx}
\usepackage{booktabs}

\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
\documentclass[11pt]{book}

\begin{document}

\title{Solving the Traveling Salesman Problem}
\author{Brennan Hoeting}
\date{}
\maketitle

\section{The problem}
The traveling salesman problem can be described as follows:
given a list of cities, find the route that visits each city
exactly once and returns to the starting city with the least possible total
distance traveled.

\end{document}

### 2017-04-29 bruteforce/tsp.py
rename graph

@@ -89,11 +89,11 @@ def main():
         points[i] = all_points[i]
 
     # Generate TSP graph
-    G = find_tsp(points)
+    graph = find_tsp(points)
 
     # Draw it
-    nx.draw(G,
-            nx.get_node_attributes(G, 'pos'),
+    nx.draw(graph,
+            nx.get_node_attributes(graph, 'pos'),
             with_labels=False,
             node_size=20)
 

### 2017-04-29 bruteforce/tsp.py
update bruteforce to return edge matrix

@@ -4,6 +4,7 @@ import random
 import itertools
 import networkx as nx
 import matplotlib.pyplot as plt
+from pandas import *
 
 
 def read_tsp(filename):
@@ -23,79 +24,139 @@ def read_tsp(filename):
     return points
 
 
+def create_distance_matrix(points):
+    """
+    Create a matrix from a dict of points
+    where matrix[i, j] is the distance
+    between point i and point j, where
+    i and j also corrospond to the keys 
+    in points
+    """
+    n = len(points) + 1
+
+    matrix = [[0]*n for _ in range(n)]
+
+    for i, point1 in points.items():
+        for j, point2 in points.items():
+            if i < j:
+                matrix[i][j] = find_distance(points[i], points[j])
+                matrix[j][i] = matrix[i][j]
+
+    return matrix
+
+
+def print_matrix(matrix):
+    print(DataFrame(matrix))
+
+
 def find_distance(point1, point2):
     return math.sqrt(
             ((point2[1] - point1[1]) ** 2) +
             ((point2[0] - point1[0]) ** 2))
 
 
-def path_distance(path, points):
-    distance = find_distance(points[path[-1]], points[path[0]])
-    for index in range(len(path)):
-        if index == len(path) - 1:
-            break
-        distance += find_distance(points[path[index]], points[path[index+1]])
+def find_path_distance(path, matrix):
+    distance = 0
+    for i in range(1, len(path)):
+        distance += matrix[path[i-1]][path[i]]
 
     return distance
 
 
-def find_tsp(points):
-    # Grab the starting key and remove the
-    # starting point from the points dict
-    starting_index = next(iter(points.keys()))
-    destination_points = points.copy()
-    del destination_points[starting_index]
+def create_edge_matrix(path):
+    path_length = len(path)
+    edge_matrix = [[0]*path_length for _ in range(path_length)]
+    for i in range(1, path_length):
+        edge_matrix[path[i-1]][path[i]] = 1
+        #edge_matrix[path[i]][path[i-1]] = 1
+
+    return edge_matrix
+
+
+def find_tsp(distance_matrix):
+
+    # Use the first point as the starting point
+    # (this is not zero because the TPS dataset
+    # is indexed at 1.
+    start = 1
+    n = len(distance_matrix)
+
+    # The destination points will be every point
+    # except the starting point.  By "point",
+    # we man index.
+    destinations = range(start + 1, n)
 
     # Generate all the permutations of the destination points.
-    # The starting point will be inserted at the beginnging of
[…]

### 2017-04-29 bruteforce/tsp.py
python needs stricter scoping

@@ -111,7 +111,7 @@ def find_tsp(distance_matrix):
     # another matrix where matrix[i][j] will
     # either be 1 if there is an edge between
     # i and j, or 0 otherwise.
-    return create_edge_matrix(path)
+    return create_edge_matrix(shortest_path)
     
 
 def create_graph(edge_matrix, points):

### 2017-04-29 paper/paper.tex
update bruteforce to return edge matrix
add abstract and intro

@@
 \usepackage{mathptmx}
-\usepackage{booktabs}
 
-\DeclarePairedDelimiter\ceil{\lceil}{\rceil}
-\DeclarePairedDelimiter\floor{\lfloor}{\rfloor}
-\documentclass[11pt]{book}
+\documentclass[12pt]{report}
 
@@
 
-\title{Solving the Traveling Salesman Problem}
+\title{Exploring the Traveling Salesman Problem}
 \author{Brennan Hoeting}
@@
 
-\section{The problem}
+\section*{Abstract}
+This paper will explore the Traveling Salesman Problem.
+We will first obtain an understanding of the problem, that
+is, we will explain the problem, examine its input and output,
+and identify what makes solving this problem particularly
+challenging.  Next, we will present a naive brute force solution
+to the problem and analyze it's efficiency and correctness.  We
+will improve upon that solution by comparing and contrasting
+more efficient approaches to solving the problem.  We will choose
+an efficient solution to present, explain, and compare runtimes
+against naive solution.  We will prove that our efficient solution
+can solve a subclass of instances of the Traveling Salesman Problem.
+Finally, Python code for the naive and efficient solutions will be
+presented and demonstrated.
+
+\section*{Understanding the Problem}
 The traveling salesman problem can be described as follows:
-given a list of cities, find the route that visits each city
-exactly once and returns to the starting city with the least possible total
-distance traveled.
+Given a set of cities, where there is a distance between
+each pair of cities, find a tour one could take that 
+begins at a starting city, visits each destination city
+exactly once, and returns to the starting city with the
+minimum possible total distance traveled. \par
+
+The problem input is a matrix that defines the distance
+between each pair of cities.  We will denote this as
+the distance matrix $D_{ij}$, where
+
+\section*{A Naive Brute Force Solution}
 

### 2017-04-29 bruteforce/tsp.py
use diagonal matricies in bf

@@ -40,7 +40,6 @@ def create_distance_matrix(points):
         for j, point2 in points.items():
             if i < j:
                 matrix[i][j] = find_distance(points[i], points[j])
-                matrix[j][i] = matrix[i][j]
 
     return matrix
 
@@ -58,7 +57,9 @@ def find_distance(point1, point2):
 def find_path_distance(path, matrix):
     distance = 0
     for i in range(1, len(path)):
-        distance += matrix[path[i-1]][path[i]]
+        point1 = min(path[i-1], path[i])
+        point2 = max(path[i-1], path[i])
+        distance += matrix[point1][point2]
 
     return distance
 
@@ -67,8 +68,9 @@ def create_edge_matrix(path):
     path_length = len(path)
     edge_matrix = [[0]*path_length for _ in range(path_length)]
     for i in range(1, path_length):
-        edge_matrix[path[i-1]][path[i]] = 1
-        #edge_matrix[path[i]][path[i-1]] = 1
+        point1 = min(path[i-1], path[i])
+        point2 = max(path[i-1], path[i])
+        edge_matrix[point1][point2] = 1
 
     return edge_matrix
 
@@ -147,7 +149,6 @@ def main():
  
     # Create the edge matrix
     edge_matrix = find_tsp(matrix)
-    print_matrix(edge_matrix)
 
     # Create the graph
     graph = create_graph(edge_matrix, points)

### 2017-04-30 paper/paper.tex
finish bruteforce

@@
+%\documentclass[margin, 11pt]{article}
 \documentclass{article}
+\usepackage[left=3cm,right=3cm,top=2cm,bottom=2cm]{geometry}
 
-\usepackage[left=3cm,right=3cm,top=2cm,bottom=2cm]{geometry}
-\usepackage[fleqn]{amsmath}
+\usepackage{amsmath}
 \usepackage{amssymb}
@@
 \usepackage{clrscode3e}
+\usepackage{blkarray}
+\usepackage{booktabs}
 \usepackage{parskip}
 \usepackage{mathptmx}
-
-\documentclass[12pt]{report}
+\usepackage{bigstrut}
+\usepackage{mathptmx}
 
@@
 
-\section*{Abstract}
-This paper will explore the Traveling Salesman Problem.
+\section*{Introduction}
+This paper will explore the Traveling Salesman Problem (TSP).
 We will first obtain an understanding of the problem, that
@@
 exactly once, and returns to the starting city with the
-minimum possible total distance traveled. \par
+minimum possible total distance traveled.  We will refer
+to the tour with the minimum possible distance traveled
+as the \textit{optimal tour}.
+\par
 
@@
 between each pair of cities.  We will denote this as
-the distance matrix $D_{ij}$, where
+the distance matrix $D = (d_{ij})$, where $d_{ij}$ is
+the distance between city $i$ and city $j$.  It is important
+to note that the distance from city $i$ to city $j$ is equal
+to the distance from city $j$ to city $i$. This means that our 
+TSP variation is \textit{undirected} so we may use a diagnol 
+matrix to represent the distances.  That is, for every $d_{ij}$ 
+where $i<j$, $d_{ij}$ is the distance between cities $i$ and $j$,
+and every $d_{ij}$ where $i\geq j$, $d_{ij} = 0$.  In a \textit{directed}
+variation of the TSP --- such as one that optimizes for minimum total
+airfair to travel the tour by plane, where ticket prices from city
+$i$ to city $j$ is not necessarily equal to the ticket price from 
+city $j$ to city $i$ --- we could not store the costs in a diagnol
+matrix.  See below for a valid input example of our TSP variation.
+Notice, for example, that the distance between city $1$ and city $4$
+is $d_{1,4}=29$.  Observe that when $i<j$ $d_{ij}=0$, implying
+we will ignore distance between $i$ and $j$, either because $i=j$
+(meaning we don't travel from city $i$ to city $i$) or because $j>i$,
+in which case the distance can be found elsewhere in the matrix.
+\par
 
-\section*{A Naive Brute Force Solution}
+\begin{align*}
+  d_{ij}=
+  \begin{blockarray}{cccccc}
+    & j_1 & j_2 & j_3 & j_4 & j_5 \\
+  \begin{block}{c(ccccc)}
+    i_1 & \cdot & 7 & 37 & 29 & 17 \\
+    i_2 & \cdot & \cdot & 31 & 32 & 34 \\
+    i_3 & \cdot & \cdot & \cdot & 32 & 42 \\
+    i_4 & \cdot & \cdot & \cdot & \cdot & 20 \\
+    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
+  \end{block}
+  \end{blockarray}
+\end{align*}
+
+Similarly, the output for the TSP is also a matrix with the dimensions
+$N\times N$, where $N$ is the number of cities.  We will define this as
+the edge matrix $E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
+going from city $i$ to city $j$.  Otherwise, $e_{ij}=0$, meaning we will not
+travel from city $i$ to city $j$.  We call this an edge matrix because if
+we had a graph $G=(V,E)$, $V_i$ would represent city $i$ and $E_{ij}$ would
+represent traveling from city $i$ to city $j$ as part of our optimal tour.
+If the graph where visualized, each verticy would be positioned at its
+longitude and latitude coordinates, and an edge would connect each pair of
+verticies $v_i$ and $v_j$ where $e_{ij}=1$.  We will see this visual
+representation of the optimal tour as part of the code demonstration.
+Below is an example of valid output for our TSP variation.  Notice that
+$e_{3,4}=1$ implying our optimal tour entails traveling from city $3$ to
+city $4$.  Observe that when $i<j$, just as the distance matrix $d_{ij}$,
+the value in $e_{ij}$ is not used, resulting in another diagonal matrix.
+\par
+
+\begin{align*}
+  e_{ij}=
+  \begin{blockarray}{cccccc}
+    & j_1 & j_2 & j_3 & j_4 & j_5 \\
+  \begin{block}{c(ccccc)}
+    i_1 & \cdot & 1 & 0 & 0 & 1 \\ 
+    i_2 & \cdot & \cdot & 1 & 0 & 0 \\
+    i_3 & \cdot & \cdot & \cdot & 1 & 0 \\
+    i_4 & \cdot & \cdot & \cdot & \cdot & 1 \\
+    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
+  \end{block}
+  \end{blockarray}
+\end{align*}
+
+Now that the Traveling Salesman Problem is defined, we shall proceed by
+exploring different methods for arriving at a solution.
+
+\section*{A Naive Brute Force Algorithm}
+The brute force method for solving the Traveling Salesman Problem has exactly 
+two redeeming qualities: (1) it is quite simple, making implementation
+trivial, and (2) it will produce the correct tour in every TSP instance.
+Unfortunately, for instances that contain more than roughly 10 cities, the
+algorithm does not finish execution.  To see why this is the case, we will
+analyze the psuedocode below:
+\par
+\vspace{0.5cm}
+
+\begin{codebox}
+\Procname{$\proc{FIND-OPTIMAL-TOUR}(D:$ distance matrix$)$}
+\li $\id{T} \gets$ all permutations of $[2 \twodots \attrib{D}{length}]$
+\li $\id{min-dist} \gets \infty$
+\li $\id{min-tour} \gets \const{nil}$
+\li \For each tour $t \in T$
+\li   \Do 
+	$\id{t} \gets [1] + T + [1]$
+\li       $\id{dist} \gets \proc{TOUR-DISTANCE}(D, t)$
+\li	  \If $dist < \id{min-dist}$
+\li	    \Then
+              $\id{min-dist} \gets dist$
+\li	      $\id{min-tour} \gets tour$
+	    \End
+       \End
+
+\li   \Return $\proc{MAKE-EDGE-MATRIX}(t)$
+\end{codebox}
+\vspace{0.5cm}
+
+On line $1$, we generate a list containing each permutation of
+$[2 \twodots \attrib{D}{length}]$.  Our starting point is at city
+$1$, so we omit it when generating the permutations since it should
+be the first and last element of each possible tour.  On line $5$, 
+we insert the starting point $1$ in the front and back of the tour
+before calculating the tour distance.  Then we simply update $\id{min-tour}$
+and $\id{min-dist}$ as we encounter a more optimal tour.  To understand
+how the runtime of this algorithm grows as the problem size increases
+(i.e.\ as the number of cities increases), refer to the table below:
+\par
+\vspace{0.25cm}
+
+\bgroup
+\def\arraystretch{1.00}
+\begin{tabular}{c|r}
+  \# of cities $n$ & \# of possible tours $(n-1){!}$ \\
+  \hline
+  \begin{tabular}{c}
+    4 \\
+    5 \\
+    6 \\
+    7 \\
+    8 \\
+    9 \\
+    10 \\
+    11 \\
+    12
+  \end {tabular}
+  &
+  \begin{tabular}{r@{\;{=}\;}l}
+    $( 4-1){!}$ & $6$ \\
+    $(5-1){!}$ & $24$ \\
+    $(6-1){!}$ & $120$ \\
+    $(7-1){!}$ & $720$ \\
+    $(8-1){!}$ & $5040$ \\
+    $(9-1){!}$ & $40,320$ \\
+    $(10-1){!}$ & $362,880$ \\
+    $(11-1){!}$ & $3,628,800$ \\
+    $(12-1){!}$ & $39,916,800$
+  \end{tabular}
+
+\end{tabular}
+\egroup
+
+Since the runtime of this algorithm is bounded
+by the number of possible tours, we can say it has
+an upper bound of $O(n!)$.  This is unsurprising
+for a brute force algorithm.  We will now explore
+a more efficient solution to the Traveling Salesman
+Problem so we may solve instances with more than 12
+cities in a reasonable runtime.
+
+\section*{A More Thoughtful Solution}
 

### 2017-04-30 bruteforce/tsp.py
add held-karp

@@ -1,10 +1,10 @@
 import sys
 import math
-import random
-import itertools
+import operator
+from itertools import combinations, permutations, chain 
 import networkx as nx
 import matplotlib.pyplot as plt
-from pandas import *
+from pandas import DataFrame
 
 
 def read_tsp(filename):
@@ -22,6 +22,12 @@ def read_tsp(filename):
             elif "NODE_COORD_SECTION" in line:
                 reading_points = True
     return points
+
+
+def powerset(iterable):
+    "powerset([1,2,3]) --> () (1,) (2,) (3,) (1,2) (1,3) (2,3) (1,2,3)"
+    s = list(iterable)
+    return chain.from_iterable(combinations(s, r) for r in range(len(s)+1))
 
 
 def create_distance_matrix(points):
@@ -75,7 +81,7 @@ def create_edge_matrix(path):
     return edge_matrix
 
 
-def find_tsp(distance_matrix):
+def brute_force(distance_matrix):
 
     # Use the first point as the starting point
     # (this is not zero because the TPS dataset
@@ -92,7 +98,7 @@ def find_tsp(distance_matrix):
     # It's important to note that the starting index isn't
     # included, as it should be at the start of every path.
     # This will be (n-1)! possible paths.
-    paths = itertools.permutations(destinations)
+    paths = permutations(destinations)
 
     # Find the path with the least distance.
     shortest_path = None
@@ -109,19 +115,76 @@ def find_tsp(distance_matrix):
             shortest_path = path
             shortest_distance = distance
 
-    # Using the shortest path, we will create
-    # another matrix where matrix[i][j] will
-    # either be 1 if there is an edge between
-    # i and j, or 0 otherwise.
-    return create_edge_matrix(shortest_path)
+    return shortest_path
+
     
+def held_karp(distance_matrix):
+    cache = {}
+    n = len(distance_matrix) - 1
+
+    # Distance helper function
+    def dist(i, j):
+        return distance_matrix[min(i, j)][max(i, j)]
+
+    print_matrix(distance_matrix)
+
+    def print_cache():
+        print()
+        for key, value in reversed(list(cache.items())):
+            print(key, "=>", value)
+        print()
+
+    # Initialize ranges
+    subset_sizes = range(1, n-1)
+    destinations = range(2, n+1)
+
+    # Add the base cases to the cache
+    for d in destinations:
+        cache[(d, ())] = (distance_matrix[1][d], (1, ()))
+
+    # Build the cache for each set with length > 0
[…]

### 2017-05-01 bruteforce/tsp.py
fix harpkarp algorithm

@@ -1,4 +1,3 @@
-import sys
 import math
 import operator
 from itertools import combinations, permutations, chain 
@@ -7,7 +6,7 @@ import matplotlib.pyplot as plt
 from pandas import DataFrame
 
 
-def read_tsp(filename):
+def read_tsp(filename, size):
     points = {}
     reading_points = False
 
@@ -18,6 +17,8 @@ def read_tsp(filename):
                 break
             elif reading_points:
                 parts = line.split(' ')
+                if int(parts[0]) > size:
+                    break
                 points[int(parts[0])] = ((int(parts[1]), int(parts[2])))
             elif "NODE_COORD_SECTION" in line:
                 reading_points = True
@@ -82,40 +83,22 @@ def create_edge_matrix(path):
 
 
 def brute_force(distance_matrix):
-
-    # Use the first point as the starting point
-    # (this is not zero because the TPS dataset
-    # is indexed at 1.
     start = 1
     n = len(distance_matrix)
 
-    # The destination points will be every point
-    # except the starting point.  By "point",
-    # we man index.
-    destinations = range(start + 1, n)
+    print("bf n", n)
 
-    # Generate all the permutations of the destination points.
-    # It's important to note that the starting index isn't
-    # included, as it should be at the start of every path.
-    # This will be (n-1)! possible paths.
+    destinations = range(start + 1, n)
     paths = permutations(destinations)
 
-    # Find the path with the least distance.
-    shortest_path = None
-    shortest_distance = sys.maxsize
+    candidates = []
     for path in paths:
-        # Complete the route by add the starting point
-        # to the beginning and end of the path.
         path = (start,) + path + (start,)
+        candidates.append((find_path_distance(path, distance_matrix), path))
 
-        # If the path is shorter than all other
-        # test paths, set it as the shortest path.
-        distance = find_path_distance(path, distance_matrix)
-        if distance < shortest_distance:
-            shortest_path = path
-            shortest_distance = distance
-
-    return shortest_path
+    solution = min(candidates, key=operator.itemgetter(0))
+    print("bfdistance", solution[0])
+    return solution[1]
 
     
 def held_karp(distance_matrix):
@@ -138,6 +121,11 @@ def held_karp(distance_matrix):
     subset_sizes = range(1, n-1)
     destinations = range(2, n+1)
 
[…]

### 2017-05-01 bruteforce/tsp.py
cleanup

@@ -1,12 +1,17 @@
 import math
 import operator
-from itertools import combinations, permutations, chain 
 import networkx as nx
-import matplotlib.pyplot as plt
 from pandas import DataFrame
+import matplotlib.pyplot as plt
+from itertools import combinations, permutations
 
 
-def read_tsp(filename, size):
+def read_tsp_instance(filename, size):
+    """
+    Read the data from a standard TSP instance.
+    The result is a dictionary that maps an index 
+    to a tuple representing the node position.
+    """
     points = {}
     reading_points = False
 
@@ -25,12 +30,6 @@ def read_tsp(filename, size):
     return points
 
 
-def powerset(iterable):
-    "powerset([1,2,3]) --> () (1,) (2,) (3,) (1,2) (1,3) (2,3) (1,2,3)"
-    s = list(iterable)
-    return chain.from_iterable(combinations(s, r) for r in range(len(s)+1))
-
-
 def create_distance_matrix(points):
     """
     Create a matrix from a dict of points
@@ -46,22 +45,32 @@ def create_distance_matrix(points):
     for i, point1 in points.items():
         for j, point2 in points.items():
             if i < j:
-                matrix[i][j] = find_distance(points[i], points[j])
+                matrix[i][j] = calc_distance(points[i], points[j])
 
     return matrix
 
 
 def print_matrix(matrix):
+    """
+    Print a matrix using the DataFrame
+    class from the pandas library.
+    """
     print(DataFrame(matrix))
 
 
-def find_distance(point1, point2):
+def calc_distance(point1, point2):
+    """
+    Calculate the distance between two points.
+    """
     return math.sqrt(
             ((point2[1] - point1[1]) ** 2) +
             ((point2[0] - point1[0]) ** 2))
 
 
-def find_path_distance(path, matrix):
+def calc_path_distance(path, matrix):
+    """
+    Calculate the total distance of a path.
+    """
     distance = 0
     for i in range(1, len(path)):
         point1 = min(path[i-1], path[i])
@@ -72,6 +81,12 @@ def find_path_distance(path, matrix):
 
 
 def create_edge_matrix(path):
+    """
+    Create edge matrix E from a path
+    such that E[i, j] is 1 if an edge
+    exists between nodes i and j, or
+    0 otherwise.
+    """
     path_length = len(path)
     edge_matrix = [[0]*path_length for _ in range(path_length)]
     for i in range(1, path_length):
[…]

### 2017-05-01 bruteforce/tsp.py
restructure code

@@ -1,240 +1,0 @@
-import math
-import operator
-import networkx as nx
-from pandas import DataFrame
-import matplotlib.pyplot as plt
-from itertools import combinations, permutations
-
-
-def read_tsp_instance(filename, size):
-    """
-    Read the data from a standard TSP instance.
-    The result is a dictionary that maps an index 
-    to a tuple representing the node position.
-    """
-    points = {}
-    reading_points = False
-
-    with open(filename) as f:
-        lines = f.readlines()
-        for line in lines:
-            if 'EOF' in line:
-                break
-            elif reading_points:
-                parts = line.split(' ')
-                if int(parts[0]) > size:
-                    break
-                points[int(parts[0])] = ((int(parts[1]), int(parts[2])))
-            elif "NODE_COORD_SECTION" in line:
-                reading_points = True
-    return points
-
-
-def create_distance_matrix(points):
-    """
-    Create a matrix from a dict of points
-    where matrix[i, j] is the distance
-    between point i and point j, where
-    i and j also corrospond to the keys 
-    in points
-    """
-    n = len(points) + 1
-
-    matrix = [[0]*n for _ in range(n)]
-
-    for i, point1 in points.items():
-        for j, point2 in points.items():
-            if i < j:
-                matrix[i][j] = calc_distance(points[i], points[j])
-
-    return matrix
-
-
-def print_matrix(matrix):
-    """
-    Print a matrix using the DataFrame
-    class from the pandas library.
-    """
-    print(DataFrame(matrix))
-
-
-def calc_distance(point1, point2):
-    """
-    Calculate the distance between two points.
-    """
-    return math.sqrt(
-            ((point2[1] - point1[1]) ** 2) +
-            ((point2[0] - point1[0]) ** 2))
-
-
-def calc_path_distance(path, matrix):
-    """
-    Calculate the total distance of a path.
-    """
-    distance = 0
-    for i in range(1, len(path)):
-        point1 = min(path[i-1], path[i])
-        point2 = max(path[i-1], path[i])
-        distance += matrix[point1][point2]
-
-    return distance
-
-
-def create_edge_matrix(path):
-    """
-    Create edge matrix E from a path
-    such that E[i, j] is 1 if an edge
-    exists between nodes i and j, or
-    0 otherwise.
-    """
-    path_length = len(path)
-    edge_matrix = [[0]*path_length for _ in range(path_length)]
-    for i in range(1, path_length):
[…]

### 2017-05-01 src/brute_force.py

@@ -1,0 +1,25 @@
+import operator
+from common import *
+from itertools import permutations
+
+
+def brute_force(distance_matrix):
+    """
+    Solve the TSP using the brute force method. That
+    is, generate each possible path and choose
+    the one with the lowest total distance.
+    """
+    start = 1
+    n = len(distance_matrix)
+
+    destinations = range(start + 1, n)
+    paths = permutations(destinations)
+
+    candidates = []
+    for path in paths:
+        path = (start,) + path + (start,)
+        candidates.append((calc_path_distance(path, distance_matrix), path))
+
+    solution = min(candidates, key=operator.itemgetter(0))
+    return solution[1]
+

### 2017-05-01 src/common.py

@@ -1,0 +1,130 @@
+import math
+import networkx as nx
+from pandas import DataFrame
+import matplotlib.pyplot as plt
+
+
+def read_tsp_instance(filename, size):
+    """
+    Read the data from a standard TSP instance.
+    The result is a dictionary that maps an index 
+    to a tuple representing the node position.
+    """
+    points = {}
+    reading_points = False
+
+    with open(filename) as f:
+        lines = f.readlines()
+        for line in lines:
+            if 'EOF' in line:
+                break
+            elif reading_points:
+                parts = line.split(' ')
+                if int(parts[0]) > size:
+                    break
+                points[int(parts[0])] = ((int(parts[1]), int(parts[2])))
+            elif "NODE_COORD_SECTION" in line:
+                reading_points = True
+    return points
+
+
+def create_distance_matrix(points):
+    """
+    Create a matrix from a dict of points
+    where matrix[i, j] is the distance
+    between point i and point j, where
+    i and j also corrospond to the keys 
+    in points
+    """
+    n = len(points) + 1
+
+    matrix = [[0] * n for _ in range(n)]
+
+    for i, point1 in points.items():
+        for j, point2 in points.items():
+            if i < j:
+                matrix[i][j] = calc_distance(points[i], points[j])
+
+    return matrix
+
+
+def print_matrix(matrix):
+    """
+    Print a matrix using the DataFrame
+    class from the pandas library.
+    """
+    print(DataFrame(matrix))
+
+
+def calc_distance(point1, point2):
+    """
+    Calculate the distance between two points.
+    """
+    return math.sqrt(
+        ((point2[1] - point1[1]) ** 2) +
+        ((point2[0] - point1[0]) ** 2))
+
+
+def calc_path_distance(path, matrix):
+    """
+    Calculate the total distance of a path.
+    """
+    distance = 0
+    for i in range(1, len(path)):
+        point1 = min(path[i - 1], path[i])
+        point2 = max(path[i - 1], path[i])
+        distance += matrix[point1][point2]
+
+    return distance
+
+
+def create_edge_matrix(path):
+    """
+    Create edge matrix E from a path
+    such that E[i, j] is 1 if an edge
+    exists between nodes i and j, or
+    0 otherwise.
+    """
+    path_length = len(path)
+    edge_matrix = [[0] * path_length for _ in range(path_length)]
+    for i in range(1, path_length):
[…]

### 2017-05-01 src/held_karp.py

@@ -1,0 +1,67 @@
+import operator
+from itertools import combinations
+
+
+def held_karp(distance_matrix):
+    """
+    Solve the TSP using the held_karp algorihtm.  
+    """
+    # Distance helper function. This is necessary
+    # since we have a diagonal distance matrix.
+    def dist(i, j):
+        return distance_matrix[min(i, j)][max(i, j)]
+
+    # The cache will map a path S, parent node P, and total 
+    # distance d to the optimal subpath of S - {P}, its parent
+    # node, and its distance. The last added item will contain
+    # the optimal distance, and we can backtrace through the
+    # cache to obtain the optimal path.
+    cache = {}
+    n = len(distance_matrix) - 1
+
+    # Initialize ranges
+    subset_sizes = range(1, n - 1)
+    destinations = range(2, n + 1)
+
+    # Add the base cases to the cache (i.e. the distance of
+    # each destination node to the starting node).
+    for d in destinations:
+        cache[(d, ())] = (distance_matrix[1][d], (1, ()))
+
+    # Build the cache for each set where the length > 0
+    for s in subset_sizes:
+        for d in destinations:
+            for subset in combinations(destinations, s):
+                if d in subset:
+                    continue
+                candidates = []
+                for i, test_subset in enumerate(combinations(subset, s - 1)):
+                    k = subset[s - i - 1]
+                    distance = cache[(k, test_subset)][0] + dist(d, k)
+                    candidates.append((distance, (k, test_subset)))
+                    
+                cache[(d, subset)] = min(
+                    candidates, key=operator.itemgetter(0))
+                
+    # Compute the final optimal subpath.
+    candidates = []
+    for i, subset in enumerate(combinations(destinations, n - 2)):
+        d = destinations[n - 2 - i]
+        candidates.append((cache[(d, subset)][0] + dist(1, d), (d, subset)))
+
+    # Cache the final optimal subpath. Keep track of the key
+    # in traveler.
+    traveler = (1, tuple(destinations))
+    cache[traveler] = min(candidates, key=operator.itemgetter(0))
+
+    # Backtrack through the cache, starting with traveler,
+    # to build the optimal subpath.
+    path = []
+    while traveler is not None:
+        path.append(traveler[0])
+        if traveler in cache:
+            traveler = cache[traveler][1]
+        else:
+            traveler = None
+
+    return tuple(reversed(path))

### 2017-05-01 src/main.py

@@ -1,0 +1,45 @@
+from time import time
+from common import *
+from held_karp import *
+from brute_force import *
+from termcolor import colored
+from prettytable import PrettyTable
+
+
+def main():
+    bf_limit = 11
+    sizes = range(4, 19)
+
+    t = PrettyTable(['Cities', 'Brute Force (s)', 'Held-Karp (s)', 'BF/HK'])
+    bf_elapsed_prev = 0
+
+    for size in sizes:
+        P = read_tsp_instance('test.tsp', size)
+        D = create_distance_matrix(P)    
+
+        bf_elapsed = 0
+        if size < bf_limit:
+            start_time = time()
+            brute_force(D)
+            bf_elapsed = (time() - start_time)
+            bf_elapsed_str = format(bf_elapsed, '.5f')
+        else:
+            bf_elapsed = bf_elapsed_prev * size
+            bf_elapsed_str = colored('{0:,.5f}'.format(bf_elapsed), 'yellow')
+
+        bf_elapsed_prev = bf_elapsed
+ 
+        start_time = time()
+        held_karp(D)
+        hc_elapsed = (time() - start_time)
+        
+        t.add_row([
+            size,
+            bf_elapsed_str,
+            '{0:,.5f}'.format(hc_elapsed),
+            '{0:,.5f}'.format(bf_elapsed / hc_elapsed)])
+
+    print(t)
+
+if __name__ == "__main__":
+    main()

### 2017-05-01 src/tsp.py

@@ -1,0 +1,25 @@
+from common import *
+from held_karp import *
+from brute_force import *
+
+
+def main():
+    size = 14
+
+    P = read_tsp_instance('test.tsp', size)
+    D = create_distance_matrix(P)    
+
+    D = D[:size + 1]
+    for i in range(len(D)):
+        D[i] = D[i][:size + 1]
+
+    if 1:
+        hc_path = held_karp(D)
+        test_path(hc_path, D, P) 
+    else:
+        bf_path = brute_force(D)
+        test_path(bf_path, D, P) 
+
+
+if __name__ == "__main__":
+    main()

### 2017-05-01 src/brute_force.py
setup minted to import python code

@@ -9,17 +9,15 @@ def brute_force(distance_matrix):
     is, generate each possible path and choose
     the one with the lowest total distance.
     """
-    start = 1
     n = len(distance_matrix)
-
-    destinations = range(start + 1, n)
+    destinations = range(2, n)
     paths = permutations(destinations)
 
     candidates = []
     for path in paths:
-        path = (start,) + path + (start,)
-        candidates.append((calc_path_distance(path, distance_matrix), path))
+        path = (1,) + path + (1,)
+        candidates.append((
+            calc_path_distance(path, distance_matrix),
+            path))
 
-    solution = min(candidates, key=operator.itemgetter(0))
-    return solution[1]
-
+    return min(candidates, key=operator.itemgetter(0))[1]

### 2017-05-01 src/common.py

@@ -7,7 +7,7 @@ import matplotlib.pyplot as plt
 def read_tsp_instance(filename, size):
     """
     Read the data from a standard TSP instance.
-    The result is a dictionary that maps an index 
+    The result is a dictionary that maps an index
     to a tuple representing the node position.
     """
     points = {}
@@ -33,11 +33,10 @@ def create_distance_matrix(points):
     Create a matrix from a dict of points
     where matrix[i, j] is the distance
     between point i and point j, where
-    i and j also corrospond to the keys 
+    i and j also corrospond to the keys
     in points
     """
     n = len(points) + 1
-
     matrix = [[0] * n for _ in range(n)]
 
     for i, point1 in points.items():
@@ -107,7 +106,7 @@ def create_graph(edge_matrix, points):
         for j in range(1, len(edge_matrix)):
             if edge_matrix[i][j] == 1:
                 graph.add_edge(i, j)
-            
+
     return graph
 
 
@@ -126,5 +125,3 @@ def test_path(path, distance_matrix, points):
             node_size=60)
 
     plt.show()
-
-

### 2017-05-01 paper/paper.tex
setup minted to import python code
paper progress

@@
-%\documentclass[margin, 11pt]{article}
-\documentclass{article}
-\usepackage[left=3cm,right=3cm,top=2cm,bottom=2cm]{geometry}
+% Preamble
+% ---
+\documentclass[11pt,fleqn]{article}
 
+% Packages
+% ---
+\usepackage[
+  left=3cm,
+  right=3cm,
+  top=2cm,
+  bottom=2cm]{geometry}
 \usepackage{amsmath}
@@
 \usepackage{mathptmx}
+\usepackage{minted}
 
+% Use borland colorscheme in minted.
+\usemintedstyle{borland}
+
+% Document
+% ---
 \begin{document}
@@
 \section*{Introduction}
-This paper will explore the Traveling Salesman Problem (TSP).
-We will first obtain an understanding of the problem, that
+This paper will explore the Traveling Salesman Problem (TSP),
+a classic NP---hard problem in the field of computer science.
+We will first obtain an understanding of the problem. That
 is, we will explain the problem, examine its input and output,
-and identify what makes solving this problem particularly
-challenging.  Next, we will present a naive brute force solution
-to the problem and analyze it's efficiency and correctness.  We
-will improve upon that solution by comparing and contrasting
-more efficient approaches to solving the problem.  We will choose
-an efficient solution to present, explain, and compare runtimes
-against naive solution.  We will prove that our efficient solution
-can solve a subclass of instances of the Traveling Salesman Problem.
-Finally, Python code for the naive and efficient solutions will be
-presented and demonstrated.
+and show why solving this problem particularly challenging.
+We will present and explain a naive brute force solution to the
+problem and analyze it's efficiency.  We will present a better
+method for solving the TSP, the Held---Karp algorithm, and explain
+it in detail mathematically and with psuedocode.  There is a 
+smorgasbord of algorihtms that imporve over the brute force method,
+but they all have their drawbacks and none of them run in polynommial
+time complexity.  We will examine a few of these methods in lesser
+detail than Held---Karp and look at situtations in which one algorithm
+is favorable over another.
 
-\section*{Understanding the Problem}
+\section{Understanding the Problem}
+\subsection{Overview}
 The traveling salesman problem can be described as follows:
 Given a set of cities, where there is a distance between
-each pair of cities, find a tour one could take that 
-begins at a starting city, visits each destination city
-exactly once, and returns to the starting city with the
-minimum possible total distance traveled.  We will refer
-to the tour with the minimum possible distance traveled
-as the \textit{optimal tour}.
-\par
+each pair of cities, find a tour (route) that begins at a
+starting city, visits each destination city exactly once,
+and returns to the starting city with the minimum possible
+total distance traveled.  We will refer to the tour with the
+minimum possible distance traveled as the \textit{optimal tour}.
 
+\subsection{Problem Input}
 The problem input is a matrix that defines the distance
@@
 the distance between city $i$ and city $j$.  It is important
-to note that the distance from city $i$ to city $j$ is equal
-to the distance from city $j$ to city $i$. This means that our 
-TSP variation is \textit{undirected} so we may use a diagnol 
+to note that the cost (distance) from city $i$ to city $j$ is equal
+to the cost from city $j$ to city $i$. This means that our
+TSP variation is \textit{undirected}, so we may use a diagnol
 matrix to represent the distances.  That is, for every $d_{ij}$
 where $i<j$, $d_{ij}$ is the distance between cities $i$ and $j$,
-and every $d_{ij}$ where $i\geq j$, $d_{ij} = 0$.  In a \textit{directed}
-variation of the TSP --- such as one that optimizes for minimum total
-airfair to travel the tour by plane, where ticket prices from city
-$i$ to city $j$ is not necessarily equal to the ticket price from 
-city $j$ to city $i$ --- we could not store the costs in a diagnol
-matrix.  See below for a valid input example of our TSP variation.
-Notice, for example, that the distance between city $1$ and city $4$
-is $d_{1,4}=29$.  Observe that when $i<j$ $d_{ij}=0$, implying
-we will ignore distance between $i$ and $j$, either because $i=j$
-(meaning we don't travel from city $i$ to city $i$) or because $j>i$,
-in which case the distance can be found elsewhere in the matrix.
+and every $d_{ij}$ where $i\geq j$, is ignored.
+\par
+
+In a \textit{directed} variation of the TSP --- such as one that
+optimizes for minimum total airfair to travel a tour by plane,
+where ticket prices from city $i$ to city $j$ is not necessarily
+equal to the ticket price from city $j$ to city $i$ --- we would
+store the cost in every value of $d_{ij}$ where $i\neq $j, rather
+than storing the costs in a diagnol matrix.  This distinction,
+while helpful to be aware of, minimally affects how the problem is
+solved in the methods we will explore.
+
+\par
+
+See below for a valid 
+input example of our TSP variation.  Notice, for example, that the
+distance between city $1$ and city $4$ is $d_{1,4}=29$.  If we ran
+into a case while implementing the algorithm where we are finding
+the distance between $i$ and $j$ but $j>i$, we simply get the distance
+from $d_{ji}$ instead of $d_{ij}$.
 \par
@@
 
+\subsection{Problem Output}
 Similarly, the output for the TSP is also a matrix with the dimensions
@@
 
-Now that the Traveling Salesman Problem is defined, we shall proceed by
-exploring different methods for arriving at a solution.
+Now that the Traveling Salesman Problem is defined, we can proceed by
+exploring different methods for computing the optimal solution.
 
-\section*{A Naive Brute Force Algorithm}
+\section{A Naive Brute Force Algorithm}
 The brute force method for solving the Traveling Salesman Problem has exactly
@@
 before calculating the tour distance.  Then we simply update $\id{min-tour}$
-and $\id{min-dist}$ as we encounter a more optimal tour.  To understand
-how the runtime of this algorithm grows as the problem size increases
-(i.e.\ as the number of cities increases), refer to the table below:
+and $\id{min-dist}$ as we encounter a more optimal tour.
+\par
+
+Looking at the table below, we can see how the number of potential optimal
+paths increases as the number of cities increases.
 \par
@@
 \bgroup
-\def\arraystretch{1.00}
+\def\arraystretch{1.5}
 \begin{tabular}{c|r}
-  \# of cities $n$ & \# of possible tours $(n-1){!}$ \\
+  Cities $n$ & Possible Tours $(n-1){!}$ \\
   \hline
@@
 \egroup
+\vspace{0.25cm}
 
-Since the runtime of this algorithm is bounded
-by the number of possible tours, we can say it has
-an upper bound of $O(n!)$.  This is unsurprising
-for a brute force algorithm.  We will now explore
-a more efficient solution to the Traveling Salesman
-Problem so we may solve instances with more than 12
-cities in a reasonable runtime.
+The runtime of this algorithm is bounded by the amount
+of time it takes to generate each possible tour.
+Therefore, we may informally say that the upper runtime
+bound for the runtime is $O(n!)$, where $n$ is the
+number of cities.
 
-\section*{A More Thoughtful Solution}
+\par
+
+\section{A More Thoughtful Solution}
+
+% Appendicies
+% ---
+\newpage
+\appendix
+\section{Appendices}
+\addcontentsline{toc}{section}{Appendices}
+\renewcommand{\thesubsection}{\Alph{subsection}}
+
+\subsection{Collaboration}
+  I collaborated with Grant Eaton and Luke Artnak.
+
+\newpage
+\subsection{Code Demonstration}
+  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
 

### 2017-05-01 paper/paper.tex
finish brute force section

@@
 \usepackage{minted}
+\usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
 
@@
 method for solving the TSP, the Held---Karp algorithm, and explain
-it in detail mathematically and with psuedocode.  There is a 
-smorgasbord of algorihtms that imporve over the brute force method,
-but they all have their drawbacks and none of them run in polynommial
+it in detail mathematically and with pseudocode.  There is a
+smorgasbord of algorithms that improve over the brute force method,
+but they all have their drawbacks and none of them run in a polynomial
 time complexity.  We will examine a few of these methods in lesser
-detail than Held---Karp and look at situtations in which one algorithm
+detail than Held---Karp and look at situations in which one algorithm
 is favorable over another.
@@
 each pair of cities, find a tour (route) that begins at a
-starting city, visits each destination city exactly once,
+starting city, visits each \textit{destination city} exactly once,
 and returns to the starting city with the minimum possible
@@
 equal to the ticket price from city $j$ to city $i$ --- we would
-store the cost in every value of $d_{ij}$ where $i\neq $j, rather
+store a cost in every value of $d_{ij}$ where $i\neq j$, rather
 than storing the costs in a diagnol matrix.  This distinction,
@@
 solved in the methods we will explore.
-
 \par
 
-See below for a valid 
-input example of our TSP variation.  Notice, for example, that the
-distance between city $1$ and city $4$ is $d_{1,4}=29$.  If we ran
-into a case while implementing the algorithm where we are finding
-the distance between $i$ and $j$ but $j>i$, we simply get the distance
-from $d_{ji}$ instead of $d_{ij}$.
+See below for a valid input example for an undirected TSP variation.
+Notice, for example, that the distance between city $1$ and city $4$ is
+$d_{1,4}=29$.  If we ran into a case while implementing a TSP algorithm
+where we are finding the distance between $i$ and $j$ but $j>i$, we
+simply retrieve the distance from $d_{ji}$ instead of $d_{ij}$.
 \par
@@
 \subsection{Problem Output}
-Similarly, the output for the TSP is also a matrix with the dimensions
-$N\times N$, where $N$ is the number of cities.  We will define this as
+The output for the TSP is also a matrix with the dimensions
+$N\times N$, where $N$ is the number of cities.  We will define the output as
 the edge matrix $E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
-going from city $i$ to city $j$.  Otherwise, $e_{ij}=0$, meaning we will not
-travel from city $i$ to city $j$.  We call this an edge matrix because if
-we had a graph $G=(V,E)$, $V_i$ would represent city $i$ and $E_{ij}$ would
-represent traveling from city $i$ to city $j$ as part of our optimal tour.
+traveling from city $i$ to city $j$.  $e_{ij}=0$ if meaning we will not
+travel from city $i$ to city $j$ along the optimal tour.
+\par
+
+We use an edge matrix because if we had a graph $G=(V,E)$, $V_i$
+would represent city $i$ and $E_{ij}$ would
+represent traveling from city $i$ to city $j$ along the optimal tour.
 If the graph where visualized, each verticy would be positioned at its
-longitude and latitude coordinates, and an edge would connect each pair of
-verticies $v_i$ and $v_j$ where $e_{ij}=1$.  We will see this visual
-representation of the optimal tour as part of the code demonstration.
-Below is an example of valid output for our TSP variation.  Notice that
-$e_{3,4}=1$ implying our optimal tour entails traveling from city $3$ to
+longitude and latitude coordinates and a segment would connect each pair
+of verticies $v_i$ and $v_j$ where $e_{ij}=1$.
+\par
+
+Below is an example of a valid output for our TSP variation.  Notice that
+$e_{3,4}=1$ implying our optimal tour involves traveling from city $3$ to
 city $4$.  Observe that when $i<j$, just as the distance matrix $d_{ij}$,
 the value in $e_{ij}$ is not used, resulting in another diagonal matrix.
+Observe that $e_1$ has two columns that are $1$.  This will always be the
+case for a single $e_i$ where $i$ is the city at we we begin and terminate
+the tour.  In this case, and throughout the paper, the optimal tour will
+always start and end on the first city.
 \par
@@
 
-Now that the Traveling Salesman Problem is defined, we can proceed by
-exploring different methods for computing the optimal solution.
-
 \section{A Naive Brute Force Algorithm}
+\subsection{Overview}
 The brute force method for solving the Traveling Salesman Problem has exactly
@@
 Unfortunately, for instances that contain more than roughly 10 cities, the
-algorithm does not finish execution.  To see why this is the case, we will
-analyze the psuedocode below:
+algorithm does not finish execution in a reasonable time.  Analyzing the
+pseudocode in (2.2), we can get a good idea of why this is the case.
 \par
@@
 
+\subsection{Pseudocode}
 \begin{codebox}
-\Procname{$\proc{FIND-OPTIMAL-TOUR}(D:$ distance matrix$)$}
+\Procname{$\proc{OPTIMAL-TOUR}(D:$ distance matrix$)$}
 \li $\id{T} \gets$ all permutations of $[2 \twodots \attrib{D}{length}]$
@@
 
-\li   \Return $\proc{MAKE-EDGE-MATRIX}(t)$
+\li    \Return $\proc{MAKE-EDGE-MATRIX}(\id{min-tour})$
 \end{codebox}
-\vspace{0.5cm}
 
-On line $1$, we generate a list containing each permutation of
-$[2 \twodots \attrib{D}{length}]$.  Our starting point is at city
-$1$, so we omit it when generating the permutations since it should
-be the first and last element of each possible tour.  On line $5$,
-we insert the starting point $1$ in the front and back of the tour
-before calculating the tour distance.  Then we simply update $\id{min-tour}$
-and $\id{min-dist}$ as we encounter a more optimal tour.
+On line $1$, we assign $\id{T}$ to a list containing each permutation of
+the seed list $[2 \twodots \attrib{D}{length}]$.  Each list $T_k$ represents
+a partial tour such that $T_{k,i+1}$ is visited from $T_{k,i}$.  Each element
+$T_{k,i}$ in $T_k$ corresponds to a row index $i$ in the distance matrix $D$.
 \par
 
+Notice that we initially exclude our starting city, $1$, from the seed list.
+This is to ensure $T$ only includes permutations the destination cities. Next,
+for each permutation $t$ in $T$, we append $1$ to the front and back of $t$.
+This gives us a complete tour, starting on $1$, visiting each destination
+exactly once, and ending on $1$.  Now we may calculate the total distance,
+$\id{dist}$, of $t$ by calling the $\proc{TOUR-DISTANCE}(D, t)$ procedure,
+which is simply:
+
+\vspace{.27cm}
+\hrule
+\begin{codebox}
+\Procname{$\proc{TOUR-DISTANCE}(t:$ tour, $D:$ distance matrix$)$}
+\li $\id{dist} \gets 0$
+\li \For $\id{i} \gets 2$ to $\attrib{t}{length}$
+\li   \Do
+$\id{dist} \gets \id{dist} + D_{\proc{min}(t_{i-1}, t_i), \proc{max}(t_{i-1}, t_i)}$
+       \End
+\li    \Return $\id{dist}$
+\end{codebox}
+\vspace{.1cm}
+\hrule
+\vspace{.05cm}
+
+If $\id{dist} < \id{min-dist}$, we update
+$\id{min-dist}$ and $\id{min-tour}$.  Once the loop is finished, we
+return an edge matrix to represent the optimal tour using the
+$\proc{MAKE-EDGE-MATRIX}()$ procedure, which creates an edge matrix from a tour
+represented as a list.
+\par
+
+\subsection{Runtime}
 Looking at the table below, we can see how the number of potential optimal
@@
 \par
-\vspace{0.25cm}
 
 \bgroup
-\def\arraystretch{1.5}
+\def\arraystretch{1.25}
 \begin{tabular}{c|r}
@@
   \begin{tabular}{c}
-    4 \\
-    5 \\
-    6 \\
-    7 \\
-    8 \\
-    9 \\
-    10 \\
-    11 \\
-    12
+    $4$ \\
+    $5$ \\
+    $\dots$ \\
+    $10$ \\
+    $11$ \\
+    $12$
   \end {tabular}
@@
   \begin{tabular}{r@{\;{=}\;}l}
-    $( 4-1){!}$ & $6$ \\
+    $(4-1){!}$ & $6$ \\
     $(5-1){!}$ & $24$ \\
-    $(6-1){!}$ & $120$ \\
-    $(7-1){!}$ & $720$ \\
-    $(8-1){!}$ & $5040$ \\
-    $(9-1){!}$ & $40,320$ \\
+    $\dots$ & $\dots$ \\
     $(10-1){!}$ & $362,880$ \\
@@
 \egroup
-\vspace{0.25cm}
 
-The runtime of this algorithm is bounded by the amount
-of time it takes to generate each possible tour.
-Therefore, we may informally say that the upper runtime
-bound for the runtime is $O(n!)$, where $n$ is the
+The runtime of this method is upper bounded by the amount
+of time it takes to generate each possible tour, so we can
+say the algorithm is $O((n-1)!) = O(n!)$, where $n$ is the
 number of cities.
-
 \par

### 2017-05-02 paper/paper.tex
progress check

@@
 \usepackage[pdftex,pdfpagelabels,bookmarks,hyperindex,hyperfigures]{hyperref}
+\usepackage{apacite}
+%\usepackage{natbib}
 
@@
 \begin{document}
+%\setlength{\abovedisplayskip}{0pt}
+%\setlength{\belowdisplayskip}{0pt}
+%\setlength{\abovedisplayshortskip}{0pt}
+%\setlength{\belowdisplayshortskip}{0pt}
 
@@
 
-\section{A More Thoughtful Solution}
+\section{A Better Algorithm}
+\subsection{Overview}
+While the brute force algorithm is simple and correct, it
+is far too slow for larger inputs.  Fortunately, there are multiple
+alternative TSP algorithms with a non-polynomial runtime
+that is better than $O(n!)$.  We will explore in detail the
+\textit{Held--Karp} algorithm, an $O(2^n)$ solution that uses
+dynamic programming \cite{heldkarp}.  Then we will look at it
+to the MST approximation algorithm and discuss the advantages
+of each approach.
+
+\subsection{Held--Karp}
+The Held--Karp algorithm is an $O(2^n)$ approach to solving the TSP\@.
+It is not fast enough to solve TSP instances with a large number of
+cities, but the solution for the instances it can solve is guarenteed
+to be correct.  The algorithm works as follows:
+\par
+
+Suppose $D = (d_{ij})$ is distance
+matrix where $d_{ij} = d_{ji}$, $n$ is the number of rows in $D$,
+$S$ is a subset of $\{1, 2,\dots,n\}$, and $x\in S$.  We define $C(S, x)$
+to be the distance of the optimal tour that starts at $x$, visits each
+destination exactly once, and ends at $1$ \cite{heldkarp}.  The recurrence
+for $C(S,x)$ is:
+
+\begin{align*}
+  C(S, x)=\begin{cases}
+    \min\{k\in S-\{x\} : d_{kx} + C(S-\{x\}, k) \}  & \text{if } n>1 \\
+    d_{1x}                                          & \text{if } n=1
+  \end{cases}
+\end{align*}
+
+We obtain the optimal cost from a distance matrix $D$ of size $n$ by calling
+$C(\{1,\dots,n\}, 1)$.  We can apply dynamic programming principles to this
+problem by storing the cost for each call in a lookup table $H$.  In each
+recurrence where $n>1$, if the tuple $(S, x)$ exists in $H$, we can return
+$H[(S, x)]$.  Otherwise,  we will compute the cost and store it in $H$.
+\par
+
+This alone does not solve the original problem.  We still need to return 
+the optimal tour, not just its distance.  We can find the optimal tour
+by backtracing through $H$.  We define the optimal tour as $P$ and $B(v)$
+as a recurrence that backtraces through $H$:
+
+\begin{align*}
+  B(v)=\begin{cases}
+    B(H[v])     & \text{if } v \text{\ in\ } H \\
+    \const{nil} & \text{otherwise} 
+  \end{cases}
+\end{align*}
+
+We fill in the values of $P$ by calling $B((\{1,\dots,n\}, 1))$.  On each
+recurrence where $v$ is in $H$, we append the second value from the tuple $v$
+to $P$.  When $v$ is not in $H$ and $\const{nil}$ is returned from $B(v)$, we
+append $1$ to $P$, resulting in the optimal tour.
+\par
+
+\subsection{Christofides}
+Christofide's algorithm is an $O(n^3)$ approximation algorithm that guarentees
+the cost of its solution will be strictly within $3/2$ of the optimal solution
+cost.  The algorithm only works on a symmetric distance matrix that satisfies
+the triangle inequality \cite{nicos}.  That is, a distance matrix $D=d_{ij}$
+satisfies the triangle inequality if the shortest path to reach every $j$ from
+$i$ involves traveling directly to $j$ from $i$, rather than through a path that
+includes more than two verticies.
+\par
+
+The algorithm works as follows: Given a distance matrix $D=d_{ij}$,
 
@@
 % ---
+%\newpage
+%\appendix
+%\section{Appendices}
+%\addcontentsline{toc}{section}{Appendices}
+%\renewcommand{\thesubsection}{\Alph{subsection}}
+%
+%\subsection{Collaboration}
+%  I collaborated with Grant Eaton and Luke Artnak.
+%
+%\newpage
+%\subsection{Code Demonstration}
+%  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
+%
 \newpage
-\appendix
-\section{Appendices}
-\addcontentsline{toc}{section}{Appendices}
-\renewcommand{\thesubsection}{\Alph{subsection}}
-
-\subsection{Collaboration}
-  I collaborated with Grant Eaton and Luke Artnak.
-
-\newpage
-\subsection{Code Demonstration}
-  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
+\bibliographystyle{apacite}
+\bibliography{biblio}
 

### 2017-05-02 paper/paper.tex
finish algorithm comparison

@@
 \usepackage{apacite}
-%\usepackage{natbib}
+\usepackage{tikz}
 
@@
 the triangle inequality \cite{nicos}.  That is, a distance matrix $D=d_{ij}$
-satisfies the triangle inequality if the shortest path to reach every $j$ from
+may be used if the shortest path to reach every $j$ from
 $i$ involves traveling directly to $j$ from $i$, rather than through a path that
@@
 
-The algorithm works as follows: Given a distance matrix $D=d_{ij}$,
+The algorithm works as follows: Given a distance matrix $D=d_{ij}$ and size $n$,
+let $G=(V,E)$ where $v_i=i$ for all $i\in [1,\dots,n]$ and $e_{ij}=d_{ij}$.  The
+optimal tour starting at $v_1$ is obtained using the following steps:
+\begin{enumerate}
+  \item Generate a minimum spanning tree, $MST$, rooted at $v_1$ using Prim's algorithm.
+  \item Perform a preorder traversal on $MST$ and append the verticies in the order they 
+    are visited to list a $T$.
+  \item For each verticy in $T$, append it to $T'$ only when it is first visited.  This 
+    removes duplicates from $T$ resulting in the optimal tour.
+\end{enumerate}
+
+The last step holds since $G$ satisfies the triangle inequality.
+
+\subsection{Nearest Neighbor}
+The Nearest Neighbor algorithm is a greedy approximation algorithm
+for solving the TSP\@.  It starts at an arbitrary city, then visits
+the closest city until each city has been visited once.  After each
+visit, the city is added to the optimal tour $T$.  Once each node
+has been visited, the starting city is appended to $T$ and $T$ is
+returned. \cite{bellmore}
+\par
+
+The steps for this algorithm are as follows:
+\begin{enumerate}
+  \item Pick a starting city $s$, let it be the current city $x$.
+  \item Add $x$ to the optimal tour $T$ and mark $x$ as visited.
+  \item Determine the closest unvisited city $y$ from $x$.  If no
+    unvisited cities exist, goto step 5.
+  \item assign $x$ to $y$, repeat step 2.
+  \item Add $s$ to $T$, return $T$.
+\end{enumerate}
+
+\subsection{Comparing the Three}
+We have examined one exact algorithm -- Held-Karp -- and two approximation
+algorithms --Christofides and Nearest Neighbor.  The approximation algorithms
+have the advantage of being significantly faster than the Held-Karp algorithm,
+but they aren't applicable in every case, and they don't always return the optimal
+tour for cases in which they are applicable.  The Held-Karp algorithm, on the other
+hand, will return the optimal tour in every TSP instance, provided the number of
+cities is small enough to allow the algorithm to finish execution in a reasonable
+time.
+\par
+
+Christofide's algorithm and the Nearest Neighbor algorithm both perform at an
+acceptable level (that is, they return the optimal tour enough of the time) for
+Euclidean TSP instances.  If a TSP instance is Euclidean, then it satisfies the
+triangle inequality.  This property is necessary for Christofide's algorithm to
+return the correct tour \cite{nicos} and it greatly increases the probability
+that the Nearest Neighbor algorithm will as well \cite{gutin}.
+\par
+
+For asymmetric and general symmetric TSP instances, Christofide's algorithm will
+never return the optimal tour.  The Nearest Neighbor algorithm, on the other hand,
+can find an optimal tour, but it's highly unlikely.  Additionally, for every $n\gte 2$
+there exists at least one asymmetric or general symmetric TSP instance for which the
+Nearest Neighbor algorithm returns the \textit{worst} tour. \cite{gutin}.  We are
+optimizing for correctness over speed and problem size in this paper and will continue
+to explore the Held--Karp algorithm.
+
+\subsection{Held-Karp Pseudocode}
 

### 2017-05-03 paper/paper.tex
update

@@
 \usepackage{amssymb}
+\usepackage{amsthm}
 \usepackage{mathtools}
@@
 
+% Replace outlined proof square with a black square.
+\renewcommand\qedsymbol{$\blacksquare$}
+
 % Document
@@
 \begin{document}
-%\setlength{\abovedisplayskip}{0pt}
-%\setlength{\belowdisplayskip}{0pt}
-%\setlength{\abovedisplayshortskip}{0pt}
-%\setlength{\belowdisplayshortskip}{0pt}
 
+% Title
+% ---
 \title{Exploring the Traveling Salesman Problem}
@@
 
+% Introduction
+% ---
 \section*{Introduction}
-This paper will explore the Traveling Salesman Problem (TSP),
-a classic NP---hard problem in the field of computer science.
-We will first obtain an understanding of the problem. That
-is, we will explain the problem, examine its input and output,
-and show why solving this problem particularly challenging.
-We will present and explain a naive brute force solution to the
-problem and analyze it's efficiency.  We will present a better
-method for solving the TSP, the Held---Karp algorithm, and explain
-it in detail mathematically and with pseudocode.  There is a
-smorgasbord of algorithms that improve over the brute force method,
-but they all have their drawbacks and none of them run in a polynomial
-time complexity.  We will examine a few of these methods in lesser
-detail than Held---Karp and look at situations in which one algorithm
-is favorable over another.
+The Traveling Salesman Problem (TSP) is a classic NP---hard problem
+in the field of computer science.  Like every other NP--hard problem,
+it can only be solved in exponential runtime, or approximated in polynomail
+runtime.  In this paper, we will obtain an understanding of the TSP problem
+by explaining what it is, examining its input and output, and looking at
+different algorithms that can obtain or approximate a solution.  We will
+present and explain a naive brute force solution to the problem and analyze
+its efficiency.  Then we will look at alternative TSP algorithms and discuess
+the use cases and advantages of each approach.  We will choose one of the
+algorithms and present its pseudocode, as well as detailed mathematical
+notation to gain a better understanding of how it computes the solution.
 
@@
 each pair of cities, find a tour (route) that begins at a
-starting city, visits each \textit{destination city} exactly once,
-and returns to the starting city with the minimum possible
-total distance traveled.  We will refer to the tour with the
+\textit{starting city}, visits each \textit{destination city}
+exactly once, and returns to the starting city covering the minimum
+possible total distance.  We will refer to the tour with the
 minimum possible distance traveled as the \textit{optimal tour}.
@@
 The problem input is a matrix that defines the distance
-between each pair of cities.  We will denote this as
-the distance matrix $D = (d_{ij})$, where $d_{ij}$ is
-the distance between city $i$ and city $j$.  It is important
-to note that the cost (distance) from city $i$ to city $j$ is equal
-to the cost from city $j$ to city $i$. This means that our
-TSP variation is \textit{undirected}, so we may use a diagnol
-matrix to represent the distances.  That is, for every $d_{ij}$
-where $i<j$, $d_{ij}$ is the distance between cities $i$ and $j$,
-and every $d_{ij}$ where $i\geq j$, is ignored.
+between each pair of cities.  We will refer to this as
+a \textit{distance matrix} or \textit{cost matrix} and denote
+it as $D = (d_{ij})$, where $d_{ij}$ is the distance between city
+$i$ and city $j$.  If the cost (distance) from city $i$ to city $j$
+is equal to the cost from city $j$ to city $i$ then the TSP variation
+is \textit{undirected} and our distance matrix is symmetric.
 \par
 
-In a \textit{directed} variation of the TSP --- such as one that
+An example of a \textit{directed} variation of the TSP is one that
 optimizes for minimum total airfair to travel a tour by plane,
 where ticket prices from city $i$ to city $j$ is not necessarily
-equal to the ticket price from city $j$ to city $i$ --- we would
-store a cost in every value of $d_{ij}$ where $i\neq j$, rather
-than storing the costs in a diagnol matrix.  This distinction,
-while helpful to be aware of, minimally affects how the problem is
-solved in the methods we will explore.
+equal to the ticket price from city $j$ to city $i$, implying the
+cost matrix is asymmetric.
 \par
 
-See below for a valid input example for an undirected TSP variation.
-Notice, for example, that the distance between city $1$ and city $4$ is
-$d_{1,4}=29$.  If we ran into a case while implementing a TSP algorithm
-where we are finding the distance between $i$ and $j$ but $j>i$, we
-simply retrieve the distance from $d_{ji}$ instead of $d_{ij}$.
+Most exact TSP algorithms can solve both directed and undirected variations,
+while approximation TSP algorithms can typically only solve undirected
+variations.
+\par
+
+See the figure below for a valid input distance matrix for an undirected
+TSP variation.  Notice, for example, that the distance between city $1$
+and city $4$ is $d_{1,4}=d_{4,1}=29$.  If we ran into a case while
+implementing a TSP algorithm
 \par
@@
   \begin{block}{c(ccccc)}
-    i_1 & \cdot & 7 & 37 & 29 & 17 \\
-    i_2 & \cdot & \cdot & 31 & 32 & 34 \\
-    i_3 & \cdot & \cdot & \cdot & 32 & 42 \\
-    i_4 & \cdot & \cdot & \cdot & \cdot & 20 \\
-    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
+    i_1 & 0 & 7 & 37 & 29 & 17 \\
+    i_2 & 7 & 0 & 31 & 32 & 34 \\
+    i_3 & 37 & 31 & 0 & 32 & 42 \\
+    i_4 & 29 & 32 & 32 & 0 & 20 \\
+    i_5 & 17 & 34 & 32 & 20 & 0 \\
   \end{block}
@@
 
+It is common to describe the input of the TSP as a complete graph.
+We define the graph as $G=(V,E)$, where $v_i$ is the $i^{\text{th}}$
+city and $e_ij$ is the distance between city $i$ and city $j$.  Below
+is an example what such a graph could look like:
+
+\tikzstyle{vertex}=[circle, draw, minimum size=1.2cm]
+\tikzstyle{edge}=[draw=black]
+\begin{tikzpicture}[shorten >=1pt, auto, node distance=7cm]
+  \node[vertex](v1) at (6, 4) {$ v_1 $};
+  \node[vertex](v2) at (11, 2) {$ v_2 $};
+  \node[vertex](v3) at (10, -2) {$ v_3 $};
+  \node[vertex](v4) at (2, -2) {$ v_4 $};
+  \node[vertex](v5) at (1, 2) {$ v_5 $};
+
+  \draw[edge] (v1) edge node{\scriptsize 7}  (v2);
+  \draw[edge] (v1) edge node{\scriptsize 37} (v3);
+  \draw[edge] (v1) edge node{\scriptsize 29} (v4);
+  \draw[edge] (v1) edge node{\scriptsize 17} (v5);
+
+  \draw[edge] (v2) edge node{\scriptsize 31} (v3);
+  \draw[edge] (v2) edge node{\scriptsize 32} (v4);
+  \draw[edge] (v2) edge node{\scriptsize 34} (v5);
+
+  \draw[edge] (v3) edge node{\scriptsize 32} (v4);
+  \draw[edge] (v3) edge node{\scriptsize 42} (v5);
+
+  \draw[edge] (v4) edge node{\scriptsize 42} (v5);
+\end{tikzpicture}
+
 \subsection{Problem Output}
-The output for the TSP is also a matrix with the dimensions
-$N\times N$, where $N$ is the number of cities.  We will define the output as
-the edge matrix $E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
+A common output for a TSP is a matrix with the dimensions $N\times N$, where
+$N$ is the number of cities.  We can define the output as the edge matrix
+$E = (e_{ij})$, where $e_{ij}=1$ if the optimal tour involves
 traveling from city $i$ to city $j$.  $e_{ij}=0$ if meaning we will not
@@
 
-We use an edge matrix because if we had a graph $G=(V,E)$, $V_i$
-would represent city $i$ and $E_{ij}$ would
-represent traveling from city $i$ to city $j$ along the optimal tour.
-If the graph where visualized, each verticy would be positioned at its
-longitude and latitude coordinates and a segment would connect each pair
-of verticies $v_i$ and $v_j$ where $e_{ij}=1$.
-\par
-
-Below is an example of a valid output for our TSP variation.  Notice that
-$e_{3,4}=1$ implying our optimal tour involves traveling from city $3$ to
-city $4$.  Observe that when $i<j$, just as the distance matrix $d_{ij}$,
-the value in $e_{ij}$ is not used, resulting in another diagonal matrix.
-Observe that $e_1$ has two columns that are $1$.  This will always be the
-case for a single $e_i$ where $i$ is the city at we we begin and terminate
-the tour.  In this case, and throughout the paper, the optimal tour will
-always start and end on the first city.
+Below is an example of a valid edge matrix output for our TSP variation.
+Notice that $e_{3,4}=1$ implying our optimal tour involves traveling from
+city $3$ to city $4$.  Since this output is for an undirected TSP instance,
+$e_{4,3}$ is also $1$.
 \par
@@
   \begin{block}{c(ccccc)}
-    i_1 & \cdot & 1 & 0 & 0 & 1 \\
-    i_2 & \cdot & \cdot & 1 & 0 & 0 \\
-    i_3 & \cdot & \cdot & \cdot & 1 & 0 \\
-    i_4 & \cdot & \cdot & \cdot & \cdot & 1 \\
-    i_5 & \cdot & \cdot & \cdot & \cdot & \cdot \\
+    i_1 & 0 & 1 & 0 & 0 & 1 \\
+    i_2 & 1 & 0 & 1 & 0 & 0 \\
+    i_3 & 0 & 1 & 0 & 1 & 0 \\
+    i_4 & 0 & 0 & 1 & 0 & 1 \\
+    i_5 & 1 & 0 & 0 & 1 & 0 \\
   \end{block}
@@
 \end{align*}
+
+TSP algorithms can also about a graph $G=(V,E)$, where $V_i$
+represents city $i$ and $E_{ij}$ represents traveling from city $i$ to
+city $j$ along the optimal tour, where the weight is the distance between
+the two.  If the graph where visualized, each verticy would be positioned
+at its longitude and latitude coordinates and a segment would connect each pair
+of verticies $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
+represents the optimal tour.
+\par
+
+\tikzstyle{vertex}=[circle, draw, minimum size=1.2cm]
+\tikzstyle{edge}=[draw=black]
+\begin{tikzpicture}[shorten >=1pt, auto, node distance=7cm]
+  \node[vertex](v1) at (6, 4) {$ v_1 $};
+  \node[vertex](v2) at (11, 2) {$ v_2 $};
+  \node[vertex](v3) at (10, -2) {$ v_3 $};
+  \node[vertex](v4) at (2, -2) {$ v_4 $};
+  \node[vertex](v5) at (1, 2) {$ v_5 $};
+
+  \draw[edge] (v1) edge node{\scriptsize 7}  (v2);
+  \draw[edge] (v2) edge node{\scriptsize 31} (v3);
+  \draw[edge] (v3) edge node{\scriptsize 32} (v4);
+  \draw[edge] (v4) edge node{\scriptsize 42} (v5);
+  \draw[edge] (v5) edge node{\scriptsize 17} (v1);
+\end{tikzpicture}
+
+Another way to represent the optimal tour is with a list
+$T$, where for every $i \lte n$, $(i-1, i)$ represents
+traveling from city $i-1$ to city $i$ along the optimal
+tour.  This paper will primarily use this type of output.
 
@@
 trivial, and (2) it will produce the correct tour in every TSP instance.
-Unfortunately, for instances that contain more than roughly 10 cities, the
+Unfortunately, for instances that contain more than roughly 20 cities, the
 algorithm does not finish execution in a reasonable time.  Analyzing the
@@
 
-\li    \Return $\proc{MAKE-EDGE-MATRIX}(\id{min-tour})$
+\li    \Return $\id{min-tour}$
 \end{codebox}
@@
 \li   \Do
-$\id{dist} \gets \id{dist} + D_{\proc{min}(t_{i-1}, t_i), \proc{max}(t_{i-1}, t_i)}$
+$\id{dist} \gets \id{dist} + D_{i-1,i}$
        \End
@@
 $\id{min-dist}$ and $\id{min-tour}$.  Once the loop is finished, we
-return an edge matrix to represent the optimal tour using the
-$\proc{MAKE-EDGE-MATRIX}()$ procedure, which creates an edge matrix from a tour
-represented as a list.
+return the optimal tour.
 \par
@@
 is far too slow for larger inputs.  Fortunately, there are multiple
-alternative TSP algorithms with a non-polynomial runtime
+alternative TSP algorithms with a runtime
 that is better than $O(n!)$.  We will explore in detail the
-\textit{Held--Karp} algorithm, an $O(2^n)$ solution that uses
-dynamic programming \cite{heldkarp}.  Then we will look at it
-to the MST approximation algorithm and discuss the advantages
-of each approach.
+\textit{Held--Karp} algorithm, an $O(n^2 2^n)$ solution that uses
+dynamic programming \cite{heldkarp}.  Then we will look at the Nearest
+Neighbor algorithm and Christofides algorithm, both of which are polynomial
+time approximation algorithms.  We will and discuss the advantages
+and drawbacks of each approach.
 
@@
 the optimal tour, not just its distance.  We can find the optimal tour
-by backtracing through $H$.  We define the optimal tour as $P$ and $B(v)$
-as a recurrence that backtraces through $H$:
+by using an additional lookup table, $P$, that maps $(S, x)$ to the optimal
+$(S-\{x\}, k)$ at each step.  We can obtain the final optimal tour by
+backtracking through $P$.  We define the optimal tour as $T$ and $B(v)$ as a
+recurrence that backtracks through $P$:
 
@@
   B(v)=\begin{cases}
-    B(H[v])     & \text{if } v \text{\ in\ } H \\
+    B(P[v])     & \text{if } v \text{\ in\ } P \\
     \const{nil} & \text{otherwise} 
@@
 
-We fill in the values of $P$ by calling $B((\{1,\dots,n\}, 1))$.  On each
-recurrence where $v$ is in $H$, we append the second value from the tuple $v$
-to $P$.  When $v$ is not in $H$ and $\const{nil}$ is returned from $B(v)$, we
-append $1$ to $P$, resulting in the optimal tour.
+We fill in the values of $T$ by calling $B((\{1,\dots,n\}, 1))$.  On each
+recurrence where $v$ is in $P$, we append the second value in $v$ (the $k$
+value) to $T$.  When $v$ is not in $H$ and $\const{nil}$ is returned
+from $B(v)$, we append $1$ to $P$, resulting in the optimal tour.
 \par
@@
 \subsection{Held-Karp Pseudocode}
+Below is the Pseudocode for the Held-Karp algorithm.  It is an iterative
+dynamic programming implementation of the recurrence we defined in (3.2).
+It uses two lookup tables, $C$ and $P$, for computing the optimal distances
+and backtracking to obtain the optimal path, respectively.
+\par
+
+\vspace{.27cm}
+\hrule
+\begin{codebox}
+\Procname{$\proc{HELD-KARP}(D:$ distance matrix$)$}
+\li $\id{n} \gets \attrib{D}{length}$
+\li $\id{C} \gets$ a lookup table that maps $(S, x)$ to the optimal cost
+\li $\id{P} \gets$ a lookup table that maps $(S, x)$ to the optimal $(S-\{x\}, k)$
+\li
+\li \For $\id{k} \gets 2$ to $n$
+\li   \Do
+	$\id{C} \gets D_{1k}$
+\li     $\id{P} \gets (\{\}, 1)$
+      \End
+\li
+\li \For $\id{size} \gets 2$ to $n$
+\li   \Do
+	\For $\id{x} \gets 1$ to $n$
+\li	  \Do
+	    \If $size \isequal n$ and $x \neq 1$
+\li	      \Then
+		\textbf{continue}
+	      \End
+\li	    \For $\id{S} \gets \proc{SUBSETS}(\{2,\dots,n\}, size)$
+\li	      \Do
+		$\id{opt-k} \gets \const{nil}$
+\li		$\id{opt-dist} \gets \infty$
+\li		$\id{opt-subset} \gets \const{nil}$
+\li		\For $k$ in $S-\{x\}$
+\li		  \Do
+		    $\id{dist} \gets \id{C}[(S-\{x\}, \id{k})] + D_{xk}$
+\li		    \If $dist < \id{opt-dist}$
+\li		      \Then
+			$\id{opt-k} \gets k$
+\li			$\id{opt-dist} \gets dist$
+\li			$\id{opt-subset} \gets S$
+		      \End
+		  \End
+\li
+\li		  $\id{C}[(subset, x)] \gets opt-dist$
+\li		  $\id{P}[(subset, x)] \gets (opt-subset, opt-k)$
+	      \End
+	  \End
+      \End
+
+\li
+\li $T = []$
+\li $\id{v} \gets (\{1,\dots,n\}, 1)$
+\li \While $\id{v} \neq \const{nil}$
+\li   \Do
+	$\proc{\attrib{T}{append}}(v[2])$
+\li	$\id{S} \gets \id{T}[v]$
+      \End	
+\li
+\li \Return \id{T}
+\end{codebox}
+\vspace{.27cm}
+\hrule
+
+On lines 5-7, we add our base cases to the lookup tables.  In the main loop on
+lines 9-25, we build the lookup tables in a bottom-up dynamic programming fashion.
+Notice on lines 11 and 12 we continue when the subset size is $n$ and $x \neq 1$.
+This acheives the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
+ensuring we're only computing optimal paths that start at $1$.  Finally, on lines 27-31
+we backtrack through our lookup table $P$ that stores the optimal $(S-\{x\}, k)$
+for a given $(S, x)$, appending each optimal $k$ value to the optimal tour $T$.
+
+\subsection{Held-Karp Runtime}
+We will show that the Held-Karp algorithm is an improvement over the brute force
+approach by proving the runtime of Held-Karp is $O(n^{2}2^n)$, which is better than
+the runtime of brute force, $O(n!)$.  Observe from the pseudocode that our algorithm
+involves computing every possible subset of $\{1,\dots,n\}$, where $n$ is the number
+of cities.  Knowing this, we can prove the runtime by proving a set with $n$ elements
+has $2^n$ subsets using induction.
+
+\begin{proof}
+  (Induction) For our basis step, consider that an empty set $\varnothing$ has $1$
+  subset (itself).  $2^n = 2^0 = 1$, so a set with $0$ elements has $2^0$ subsets.\
+  For induction, suppose a $k$-element set has $2^k$ subsets.  We will show every
+  $(k+1)$-element subset has $2^(k+1)$ subsets.  Let $A=\{a_1,a_2,\dots,a_k, b\}$,
+  notice it contains $k+1$ elements.  We parition the subsets of $A$ into two sets
+  $B$ and $C$, where $B$ contains the subsets that contain $b$ and $C$ contains
+  the subsets that don't contain $b$.  $C$ contains every subset of $A-{b}$, so it
+  has $2^k$ elements.  By construction, it follows that $B$ must have the same number
+  of elements as $C$, so it also has $2^k$ elements.  Since $B$ and $C$ are independent
+  sets, the size of $B\union C$ is $2^k + 2^k = 2^{k+1}$.
+\end{proof}
 

### 2017-05-03 src/held_karp.py

@@ -4,7 +4,7 @@ from itertools import combinations
 
 def held_karp(distance_matrix):
     """
-    Solve the TSP using the held_karp algorihtm.  
+    Solve the TSP using the held-karp algorihtm.  
     """
     # Distance helper function. This is necessary
     # since we have a diagonal distance matrix.

### 2017-05-03 paper/paper.tex
update

@@
 \usepackage{bigstrut}
+\usepackage{caption}
 \usepackage{mathptmx}
@@
 \usepackage{tikz}
+\usepackage{float}
+\usepackage{graphicx}
 
@@
 \subsection{Held--Karp}
-The Held--Karp algorithm is an $O(2^n)$ approach to solving the TSP\@.
+The Held--Karp algorithm is an $O(n^2 2^n)$ approach to solving the TSP\@.
 It is not fast enough to solve TSP instances with a large number of
@@
 
-The last step holds since $G$ satisfies the triangle inequality.
+The last step holds since $G$ satisfies the triangle inequality. \cite{nicos}
 
@@
 
+This is an $O(n^2)$ algorithm because on each iteration the
+current city visits every other unvisited city.
+
 \subsection{Comparing the Three}
@@
 acceptable level (that is, they return the optimal tour enough of the time) for
-Euclidean TSP instances.  If a TSP instance is Euclidean, then it satisfies the
+Euclidean TSP instances \cite{gutin}.  If a TSP instance is Euclidean, then it satisfies the
 triangle inequality.  This property is necessary for Christofide's algorithm to
@@
 never return the optimal tour.  The Nearest Neighbor algorithm, on the other hand,
-can find an optimal tour, but it's highly unlikely.  Additionally, for every $n\gte 2$
+can find an optimal tour, but it's highly unlikely.  Additionally, for every $n \geq 2$
 there exists at least one asymmetric or general symmetric TSP instance for which the
@@
 approach by proving the runtime of Held-Karp is $O(n^{2}2^n)$, which is better than
-the runtime of brute force, $O(n!)$.  Observe from the pseudocode that our algorithm
-involves computing every possible subset of $\{1,\dots,n\}$, where $n$ is the number
-of cities.  Knowing this, we can prove the runtime by proving a set with $n$ elements
-has $2^n$ subsets using induction.
+the runtime of brute force, $O(n!)$.
+
+Observe from the pseudocode that our algorithm iterates from $2$ to $n$.  In that
+loop we also iterate from $1$ to $n$.  This is where the $n^2$ portion of our runtime
+comes from.  Also observe that the algorithm involves computing every possible subset
+of $\{1,\dots,n\}$, where $n$ is the number of cities.  Knowing this, we can prove the
+runtime by proving a set with $n$ elements has $2^n$ subsets using induction.
 
@@
 
-% Appendicies
-% ---
-%\newpage
-%\appendix
-%\section{Appendices}
-%\addcontentsline{toc}{section}{Appendices}
-%\renewcommand{\thesubsection}{\Alph{subsection}}
-%
-%\subsection{Collaboration}
-%  I collaborated with Grant Eaton and Luke Artnak.
-%
-%\newpage
-%\subsection{Code Demonstration}
-%  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
-%
+\section{Experimental Comparison}
+
+\subsection{Overview}
+We will do an experiment on the brute force and Held--Karp algorithms,
+running Python implementations of both algorithms and comparing their
+execution times.
+
+\subsection{Input}
+The input for both algorithms will be a standard TSPLIB sample instance, as shown
+in the table below.  The distance matrix is created using the $x$ and $y$ coordinates
+in the table and is used by both algorithms to generate the optimal tour.
+\par
+
+\begin{table}[H]
+  \centering
+  \caption*{TSPLIB xqf131 input}
+  \begin{tabular}{c|c|c}
+    \toprule
+    City & $x$ & $y$ \\
+    \midrule
+    1& 0& 13 \\
+    2& 0& 26 \\
+    3& 0& 27 \\
+    4& 0& 39 \\
+    5& 2& 0\\
+    6& 5& 13\\
+    7& 5& 19\\
+    8& 5& 25\\
+    9& 5& 31\\
+    10& 5& 37\\
+    11& 5& 43\\ 
+    12& 5& 8\\ 
+    13& 8& 0\\
+    14& 9& 10\\
+    15& 10& 10\\
+    16& 11& 10\\
+    17& 12& 10\\
+    18& 12& 5\\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+\subsection{Execution}
+We run both algorithms on the data set above and produce the runtime table
+below.  Note that the brute force times for $n>10$ are estimated as the
+algorithm takes too long to run for those problem sizes.
+\par
+
+\begin{table}[H]
+  \centering
+  \caption*{Runtime comparison}
+  \begin{tabular}{c|c|c|c}
+    \toprule
+    Cities $n$ & Brute Force (s) & Held-Karp (s) & BF/HC \\
+    \midrule
+    4   &       0.00003       &    0.00005    &      0.60748      \\
+    5   &       0.00010       &    0.00011    &      0.89353      \\
+    6   &       0.00062       &    0.00032    &      1.94113      \\
+    7   &       0.00437       &    0.00091    &      4.77795      \\
+    8   &       0.03484       &    0.00226    &      15.37989     \\
+    9   &       0.29779       &    0.00568    &      52.38270     \\ 
+    10  &       2.84006       &    0.01420    &     199.93426     \\
+    11  &       31.24068      &    0.03465    &     901.55505     \\
+    12  &      374.88817      &    0.09049    &    4,143.09304    \\
+    13  &     4,873.54623     &    0.24270    &    20,080.92283   \\
+    14  &     68,229.64716    &    0.60574    &   112,639.24419   \\
+    15  &   1,023,444.70737   &    1.44135    &   710,060.87550   \\
+    16  &   16,375,115.31784  &    3.43850    &  4,762,288.58166  \\
+    17  &  278,376,960.40329  &    8.09218    &  34,400,754.72416 \\
+    18  & 5,010,785,287.25922 &    19.04262   & 263,135,354.10685 \\
+    \bottomrule
+  \end{tabular}
+\end{table}
+
+\subsection{Output}
+The output from running the Held-Karp algorithm for 18 cities is a
+list representing the optimal path:
+$(1, 2, 3, 4, 11, 10, 9, 8, 7, 6, 14, 15, 16, 17, 18, 13, 5, 12, 1)$
+
+Our program can also produce a graph:
+
+\centering
+\includegraphics[scale=0.5]{plot.png}
+
+\newpage
+\appendix
+\section*{Appendices}
+\addcontentsline{toc}{section}{Appendices}
+\renewcommand{\thesubsection}{\Alph{subsection}}
+
+\subsection{Collaboration}
+  I collaborated with Grant Eaton and Luke Artnak.
+
+\newpage
+\subsection{Code Demonstration}
+  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
+
 \newpage

### 2017-05-11 paper/paper.tex
typo

@@
 in the field of computer science.  Like every other NP--hard problem,
-it can only be solved in exponential runtime, or approximated in polynomail
+it can only be solved in exponential runtime, or approximated in polynomial
 runtime.  In this paper, we will obtain an understanding of the TSP problem
@@
 present and explain a naive brute force solution to the problem and analyze
-its efficiency.  Then we will look at alternative TSP algorithms and discuess
+its efficiency.  Then we will look at alternative TSP algorithms and discuss
 the use cases and advantages of each approach.  We will choose one of the
@@
 An example of a \textit{directed} variation of the TSP is one that
-optimizes for minimum total airfair to travel a tour by plane,
+optimizes for minimum total airfare to travel a tour by plane,
 where ticket prices from city $i$ to city $j$ is not necessarily
@@
 
-TSP algorithms can also about a graph $G=(V,E)$, where $V_i$
+TSP algorithms can also output a graph $G=(V,E)$, where $V_i$
 represents city $i$ and $E_{ij}$ represents traveling from city $i$ to
 city $j$ along the optimal tour, where the weight is the distance between
-the two.  If the graph where visualized, each verticy would be positioned
+the two.  If the graph where visualized, each vertex would be positioned
 at its longitude and latitude coordinates and a segment would connect each pair
-of verticies $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
+of vertices $v_i$ and $v_j$ where $e_{ij}=1$.  The single path in the graph
 represents the optimal tour.
@@
 It is not fast enough to solve TSP instances with a large number of
-cities, but the solution for the instances it can solve is guarenteed
+cities, but the solution for the instances it can solve is guaranteed
 to be correct.  The algorithm works as follows:
@@
 \subsection{Christofides}
-Christofide's algorithm is an $O(n^3)$ approximation algorithm that guarentees
+Christofide's algorithm is an $O(n^3)$ approximation algorithm that guarantees
 the cost of its solution will be strictly within $3/2$ of the optimal solution
@@
 $i$ involves traveling directly to $j$ from $i$, rather than through a path that
-includes more than two verticies.
+includes more than two vertices.
 \par
@@
   \item Generate a minimum spanning tree, $MST$, rooted at $v_1$ using Prim's algorithm.
-  \item Perform a preorder traversal on $MST$ and append the verticies in the order they 
+  \item Perform a preorder traversal on $MST$ and append the vertices in the order they 
     are visited to list a $T$.
-  \item For each verticy in $T$, append it to $T'$ only when it is first visited.  This 
+  \item For each vertex in $T$, append it to $T'$ only when it is first visited.  This 
     removes duplicates from $T$ resulting in the optimal tour.
@@
 Notice on lines 11 and 12 we continue when the subset size is $n$ and $x \neq 1$.
-This acheives the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
+This achieves the effect of beginning our recurrence from (3.2) with $(\{1,\dots,n\}, 1)$,
 ensuring we're only computing optimal paths that start at $1$.  Finally, on lines 27-31
@@
   $(k+1)$-element subset has $2^(k+1)$ subsets.  Let $A=\{a_1,a_2,\dots,a_k, b\}$,
-  notice it contains $k+1$ elements.  We parition the subsets of $A$ into two sets
+  notice it contains $k+1$ elements.  We partition the subsets of $A$ into two sets
   $B$ and $C$, where $B$ contains the subsets that contain $b$ and $C$ contains
@@
 \section*{Appendices}
+\newpage
 \addcontentsline{toc}{section}{Appendices}
@@
 \subsection{Collaboration}
-  I collaborated with Grant Eaton and Luke Artnak.
+  I collaborated with on this project Grant Eaton and Luke Artnak.
+  \par
+
+  Grant helped me search for primary sources to use in this paper.
+  He also made a suggestion
+  for space optimization on my Held-Karp implementation, which
+  I agreed with, but did not implement.  I helped Grant
+  with various Latex tasks and getting Python networkx code
+  to work.
+
+  Luke helped me analyze the runtime of the Held-Karp algorithm,
+  which helped me greatly in writing the runtime proof.  He also
+  made some suggestions on how I could make my Held-Karp python code
+  more readable, which again, I did not follow through with.  I helped
+  Luke get python networkx code working, although he ended up
+  switching to JavaScript later on.
+
+  \par
 
 \newpage
-\subsection{Code Demonstration}
-  \inputminted[frame=lines,linenos=true]{python}{../src/brute_force.py}
+\subsection{Source Code}
+  Brute Force Implementation
+  \inputminted[linenos=true]{python}{../src/brute_force.py}
+  \newpage
+  Held-Karp Implementation
+  \inputminted[linenos=true]{python}{../src/held_karp.py}
+  The output of the main program running on my machine.
+  Yellow values are estimations.
+  \includegraphics[scale=0.5]{screenshot.png}
+
+\newpage
+\subsection{Requirements}
+For each of the requirements of the paper, I have given
+the section where that requirement was met.
+\begin{enumerate}
+  \item Requirement: Describe the problem.  See sections 1.1, 1.2, and 1.3.
+  \item Requirement: Explain your naive / BF implementation.  See section 2.
+  \item Requirement: Motivate a solution that improves on the naive solution.
+    See section 3.2 for better solution and 3.5 for a comparison.
+  \item Requirement: Defend your solution against prior art: All of section
+    3 contains cited work, primarily in section 3.5.
+  \item Requirement: explain your improved solution by presenting pseudocode.  See section 3.6.
+  \item Requirement: Provide a proof that your solution has some nontrivial property.  See section 3.7.
+  \item Requirement: Experimentally compare your solution to BF / Naive approaches.  See section 4.
+  \item Requirement: Do version A or B of the assignment.  I've done A by choosing the Held-Karp algorithm.
+    Section 3.5 is explicitly states this paper is concerned with correctness, rather than speed.
+\end{enumerate}
 
